Concurrent Programming with CXL

### 2020-06-16 book/paper.tex
Added some text for actors

It's not easy.  It would be easy to come up with a solution that uses a single lock
and busy waiting, but we would like to avoid busy waiting.
Introduced by Dijkstra,
a \emph{semaphore} is a synchronization primitive that fits the bill.
A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
units of a resource initially available.
When allocating a resource, a process decrements the
counter using the \texttt{P} operation.  You can think of \texttt{P} standing
for \emph{Procure}, as in procuring the resource associated with the semaphore.
The \texttt{P} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{V} operation.  You can think of \texttt{V} standing for \emph{vacate},
as in vacating the resource.

@@

We have seen solutions to the reader/writer problem and the producer/consumer
problem that do not busy-wait, but they are quite different.  It would be nice
to have a general technique for all CCS problems.
The ``Split Binary Semaphore'' (SBS) is such a technique, originally proposed by
Tony Hoare and popularized by Edsger Dijkstra.
SBS is an example of what one could call a ``baton-passing'' technique.
A process needs the baton to access shared variables.
When a process has the baton and does not need it any more,
it first checks to see if there are any processes that are waiting for the
baton and can make progress.
If so, it passes the baton directly to one of those processes.
If not, it simply gives up the baton.

@@

While Hoare monitors are conceptually simpler, the so-called
``Mesa monitor semantics'' have become more popular, adopted by all major
programming languages including C, Java, and Python.

@@

When inspecting the code, one important feature to notice is that \texttt{wait} is always
invoked within a \texttt{while} loop that checks for the condition that the process is
waiting for.  We explained why in Chapter~\ref{ch:monitors}: when a process resumes after
being notified, it is not guaranteed that the condition it was waiting for holds, even
if it did at the time of invoking \texttt{notify} or \texttt{notifyAll}.
It is \emph{imperative} that you always have a \texttt{while} loop around any invocation
of \texttt{wait}.

@@

>>>>

It's not easy.  It would be easy to come up with a solution that uses a single lock
and busy waiting, but we would like to avoid busy waiting.
Introduced by Dijkstra,
a \emph{semaphore} is a synchronization primitive that fits the bill~\cite{EWD35}.
A semaphore is essentially
a counter that can be incremented and decremented but is not allowed to go
below zero.  The semaphore counter is typically initialized to the number of
units of a resource initially available.
When allocating a resource, a process decrements the
counter using the \texttt{P} operation.  You can think of \texttt{P} standing
for \emph{Procure}, as in procuring the resource associated with the semaphore.
The \texttt{P} operation blocks the invoking process if the counter is zero.
To release the resource, a process increments the resource using the
\texttt{V} operation.  You can think of \texttt{V} standing for \emph{vacate},
as in vacating the resource.

@@

We have seen solutions to the reader/writer problem and the producer/consumer
problem that do not busy-wait, but they are quite different.  It would be nice
to have a general technique for all CCS problems.
The ``Split Binary Semaphore'' (SBS) is such a technique, originally proposed by
Tony Hoare and popularized by Edsger Dijkstra~\cite{EWD703}.
SBS is an example of what one could call a ``baton-passing'' technique.
A process needs the baton to access shared variables.
When a process has the baton and does not need it any more,
it first checks to see if there are any processes that are waiting for the
baton and can make progress.
If so, it passes the baton directly to one of those processes.
If not, it simply gives up the baton.

@@

While Hoare monitors are conceptually simpler, the so-called
``Mesa monitor semantics'' or ``Mesa condition variable'' semantics
have become more popular, adopted by all major
programming languages.

That said, few programming languages provide full support for monitors.
In Java, each object has a hidden lock \emph{and} a hidden condition variable
associated with it.
Methods declared with the \texttt{synchronized} keyword automatically
obtain the lock.  Java objects also support \texttt{wait}, \texttt{notify},
and \texttt{notifyAll}.  Java also supports explicit allocations of locks
and condition variables.
In Python, locks and condition variables must be explicitly declared.
The \texttt{with} statement makes it easy to acquire and release a lock
for a section of code.
In C and C++, support for locks and condition variables is through libraries.

@@

When inspecting the code, one important feature to notice is that \texttt{wait} is always
invoked within a \texttt{while} loop that checks for the condition that the process is
waiting for.  We explained why in Chapter~\ref{ch:monitors}: when a process resumes after
being notified, it is not guaranteed that the condition it was waiting for holds, even
if it did at the time of invoking \texttt{notify} or \texttt{notifyAll}.
It is \emph{imperative} that you always have a \texttt{while} loop around any invocation
of \texttt{wait}.
Many implementation of condition variables depend on this, and optimizations of the
code allow so-called ``spurious wakeups,'' where \texttt{wait} may sometimes return even
if it has not been notified.

@@

Some programming languages favor a completely different way of implementing
synchronization using so-called \emph{actors}~\cite{HBS73}.  Actors are
processes that have only private memory and communicate through messages.
Given that there is no shared memory in the actor model, there is no need
for critical sections.  Instead, some sequential process owns a particular
piece of data and other processes access it by sending request messages
to the process and optionally waiting for response messages.  Each process
handles one message at a time, serializing all access to the data it owns.

The actor synchronization model is popular in a variety of programming
languages, including Erlang and Scala.  Actor support is also available
through popular libraries such as Akka, which is available for various
programming languages.  In Python, Java, and C/C++,
actors can be easily emulated using threads and \emph{synchronized queues}
for messaging.  Each thread would have one such queue for receiving messages.
Dequeuing from an empty synchronized queue blocks the thread until
another thread enqueues a message on the queue.

The \texttt{synch} library supports a synchronized message queue,
similar to the \texttt{Queue} object in Python.
Its interface is as follows:
\begin{itemize}
\item \texttt{Queue()} returns a new message queue;
\item \texttt{enqueue(q, item)} add \texttt{item} to the queue pointed to by \texttt{q};
\item \texttt{dequeue(q)} waits for and returns an item on the queue pointed to by \texttt{q}.
\end{itemize}

Note that a \texttt{Queue} behaves much like a zero-initialized semaphore.
\texttt{enqueue} is much like \texttt{V}, except that it is accompanied by data.
\texttt{dequeue} is much like \texttt{P}, except that it also returns data.
Thus, synchronized queues can be considered a generalization of counting semaphores.

Figure~\ref{fig:actor} shows a solution to the producer/consumer problem using
an actor.  \texttt{pc\_actor} is essentially a matchmaker: it matches
\texttt{.produce} requests with \texttt{.consume} requests.
Sometimes it may have buffered \texttt{.consume} requests and is waiting for
\texttt{.produce} requests, and sometimes it is vice versa.
The variable \texttt{balance} specifies what the situation is: if it is positive
then it has buffered \texttt{.produce} requests;
if it is negative
then it has buffered \texttt{.consume} requests.
Note that both \texttt{balance} and \texttt{requests} are variables that are
local to the \texttt{pc\_actor} process.

Each request message is a dictionary with two fields.  One of the fields is
\texttt{.type}, which either is \texttt{.produce} or \texttt{.consume}.
In case of \texttt{.produce}, the second field is \texttt{.item} and holds
the data that is produced.
In case of \texttt{.consume}, the second field is \texttt{.queue} and holds
a pointer to the queue where the response message must be enqueued.
\texttt{.produce} messages do not get a response.

The \texttt{pc\_actor} process has two arguments.  \texttt{q} is a pointer
to the queue on which it receives messages.  \texttt{nrequests} is the total
number of requests that it will handle.  We need the latter because in CXL
all processes are required to terminate.  In a more realistic implementation,
the \texttt{pc\_actor} would be in an infinite loop awaiting requests.

The \texttt{produce} method takes two arguments: a pointer to the queue
of the \texttt{pc\_actor} and the item.  The \texttt{consume} method
takes two queue pointer arguments.  The first is a pointer to the queue
of the \texttt{pc\_actor}.  The second is a pointer to the queue of the
invoker: it is the queue to which the response must be posted.

Figure~\ref{fig:actortest} shows how this code may be used.  It spawns
\texttt{NITEMS} producer and consumer processes.  Note that in that
case the \texttt{pc\_actor} is expected to receive \texttt{2 * NITEMS}
requests.
